### 2023-05-02 paper.Rmd
editing

We define the $j$'th column ($j = 1, \ldots, P$) of the original data matrix $\mathbf{X}_{N \times P}$ as $x_{\cdot j} =  (x_{1j}, \ldots, x_{Nj})$ where the column mean---the overall mean of the $j$'th attribute---is $\overline{\mathbf{x}}_j$. Let $T$ be the number of distribution moments that should be similar across $K$ groups. For each moment $t$ ($t = 1, \ldots, T$) and for each variable $x_{\cdot j}$ ($j = 1, \ldots, P$), a new *t-moment variable* $\hat{x}_{t \cdot j} = (\hat{x}_{t1j}, \ldots, \hat{x}_{tNj})$ is defined where

@@

Here, $w_t$ is the weight associated with the $t$'th moment and $\overline{\mathbf{X}}^{(k)}_{T \times P} = \{\overline{x}_{tj}^{(k)}\}_{T \times P}$ is a centroid matrix containing the mean values for all $t$-moment variables in the $k$'th group, i.e.

@@

Thus, when setting the weights for all moments to 1, the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$ has the same form as the standard k-means criterion $\mathit{SSE}$, but it is computed on different data. Computing the k-plus criterion is then done by first creating the additional $t$-moment variables, appending them as new columns to the original data matrix, and computing the $\mathit{SSE}$ on the augmented data. Using the unweighted sum approach, k-plus anticlustering therefore leaves the original k-means criterion unchanged.

Note that a problem may arise if the weights $\mathbf{w}$ are ignored when computing $\mathit{SSE_{kplusGeneralized}}$. Because the computation of the new $t$-moment variables involves exponentiation, their range of values is expected to be very different from the original variables. This imbalance can strongly and unpredictably influence the relative weight of the different moments when computing $\mathit{SSE_{kplusGeneralized}}$ (e.g., see Figure 1). To ensure a comparable weight of all criteria, all attributes---i.e., the original data as well as the new $t$-moment variables---can be standardized before computing $\mathit{SSE_{kplusGeneralized}}$, e.g. via $z$-standardization. This approach will henceforth be referred to as the unweighted standardization approach for k-plus anticlustering.

@@

To thoroughly evaluate the k-plus approach to anticlustering, two specific k-plus criteria were compared to two traditional anticlustering criteria: (a) "standard" k-plus anticlustering, minimizing differences with regard to group means and variances (by maximizing $\mathit{SSE_{kplus}}$); (b) "generalized" k-plus anticlustering, minimizing differences with regard to group means, variances, skewness and kurtosis (by maximizing $\mathit{SSE_{kplusGeneralized}}$ with $T = 4$). To implement the k-plus objectives, the unweighted standardization approach was employed that performed best in Simulation 1. Thus, to optimize a k-plus criterion, the original data set was augmented to incorporate the additional criteria and all variables---the original and the augmented variables---were $z$-standardized. Subsequently, the original k-means criterion $\mathit{SSE}$ was maximized on the basis of the augmented data set. The k-plus methods were compared to (i) k-means anticlustering, minimizing differences with regard to group means, (ii) diversity anticlustering, maximizing the sum of pairwise Euclidean distances within groups, (iii) a random allocation of elements to groups.

@@

[^dispersion]: An exception is given when using the dispersion objective, which ensures that the most similar elements are assigned to different clusters. However, anticlustering objectives striving for overall between-group similarity are mostly unaffected by the grouping of individual item pairs.

An interesting mixture of cluster analysis and anticlustering is fair clustering [e.g. @chhabra2021; @delBarrio2022]. Fair clustering performs a cluster analysis under the restriction that the distribution of a set of "protected" attributes---such as race, sex or age---should be similar between clusters. Thus, as compared to pure cluster analysis, fair clustering biases the identification of groups with regard to the protected attributes, striving for algorithmic fairness. A relevant difference to anticlustering is that fair clustering is still concerned with identifying "true" structure in the data [@delBarrio2022], while anticlustering divides data sets into artificial subgroups in such a way that between-group similarity / within-group heterogeneity is maximized. However, these methods are logically similar because anticlustering corresponds to fair clustering when only considering the protected attributes. Hence, a fruitful outlook for research might be to investigate if approaches stemming from the literature on fair clustering can be adopted for anticlustering, and vice versa.

>>>>

We define the $j$'th column ($j = 1, \ldots, P$) of the original data matrix $\mathbf{X}_{N \times P}$ as $x_{\cdot j} =  (x_{1j}, \ldots, x_{Nj})$ where the column mean---the overall mean of the $j$'th attribute---is $\overline{\mathbf{x}}_j$. Let $T$ be the number of distribution moments that should be similar across $K$ groups. For each moment $t$ ($t = 1, \ldots, T$) and for each variable $x_{\cdot j}$ ($j = 1, \ldots, P$), a new *k-plus variable* $\hat{x}_{t \cdot j} = (\hat{x}_{t1j}, \ldots, \hat{x}_{tNj})$ is defined where

@@

Here, $w_t$ is the weight associated with the $t$'th moment and $\overline{\mathbf{X}}^{(k)}_{T \times P} = \{\overline{x}_{tj}^{(k)}\}_{T \times P}$ is a centroid matrix containing the mean values for all k-plus variables in the $k$'th group, i.e.

@@

Thus, when setting the weights for all moments to 1, the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$ has the same form as the standard k-means criterion $\mathit{SSE}$, but it is computed on different data. Computing the k-plus criterion is then done by first creating the additional k-plus variables, appending them as new columns to the original data matrix, and computing the $\mathit{SSE}$ on the augmented data. Using the unweighted sum approach, k-plus anticlustering therefore leaves the original k-means criterion unchanged.

Note that a problem may arise if the weights $\mathbf{w}$ are ignored when computing $\mathit{SSE_{kplusGeneralized}}$. Because the computation of the new k-plus variables involves exponentiation, their range of values is expected to be very different from the original variables. This imbalance can strongly and unpredictably influence the relative weight of the different moments when computing $\mathit{SSE_{kplusGeneralized}}$ (e.g., see Figure 1). To ensure a comparable weight of all criteria, all attributes---i.e., the original data as well as the new k-plus variables---can be standardized before computing $\mathit{SSE_{kplusGeneralized}}$, e.g. via $z$-standardization. This approach will henceforth be referred to as the unweighted standardization approach for k-plus anticlustering.

@@

In Simulation 2, the performance of two k-plus criteria was compared to the performance of two traditional anticlustering criteria. "Standard" k-plus anticlustering was implemented by by maximizing $\mathit{SSE_{kplus}}$, i.e., minimizing differences with regard to group means and variances; "generalized" k-plus anticlustering was implemented by by maximizing $\mathit{SSE_{kplusGeneralized}}$ using $T = 4$, i.e., minimizing differences with regard to group means, variances, skewness and kurtosis. To implement the k-plus objectives, the unweighted standardization approach was employed that performed best in Simulation 1. Thus, to optimize a k-plus criterion, the original data set was augmented to incorporate the additional criteria and all variables---the original and the augmented k-plus variables---were $z$-standardized. Subsequently, the original k-means criterion $\mathit{SSE}$ was maximized on the basis of the augmented data set. The k-plus methods were compared to (i) k-means anticlustering, minimizing differences with regard to group means, (ii) diversity anticlustering, maximizing the sum of pairwise Euclidean distances within groups, (iii) a random allocation of elements to groups.

@@

[^dispersion]: An exception is given when employing the dispersion objective, which ensures that the most similar elements are assigned to different clusters. However, anticlustering objectives striving for overall between-group similarity are largely unaffected by the grouping of individual item pairs.

An interesting mixture of cluster analysis and anticlustering is fair clustering [e.g. @chhabra2021; @delBarrio2022]. Fair clustering performs a cluster analysis under the restriction that the distribution of a set of "protected" attributes---such as race, sex or age---should be similar between clusters. Thus, as compared to pure cluster analysis, fair clustering biases the group identification with regard to the protected attributes, striving for algorithmic fairness. A relevant difference to anticlustering is that fair clustering is still concerned with identifying "true" structure in the data [@delBarrio2022], while anticlustering divides data sets into artificial subgroups in such a way that between-group similarity / within-group heterogeneity is maximized. However, these methods are logically similar because fair  clustering is akin to anticlustering when only considering the protected attributes. Hence, a fruitful outlook for future research might be to investigate if approaches stemming from the literature on fair clustering can be adopted for anticlustering, and vice versa.
